---
title: "Conditional Means Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

# Practice: Conditional Means


## Motivating Example
Suppose you are a high school counselor trying to predict student GPAs as a function of other student characteristics. 

We will use the High School Longitudinal Data to answer these questions. 

First, load the data and packages. You will need ds, tidyverse, yardstick, and the hsls_clean data. 

```{r load_data, exercise=TRUE}
library(ds)
library(tidyverse)
library(yardstick)
data("hsls_clean")
```

## Unconditional Means 
If the only information we had about students was their GPA, our best guess for a given student's GPA would be the school's average GPA. 

Calculate the average GPA, use the mutate command to make a new variable for the prediction:  

```{r unconditional, exercise=TRUE, exercise.eval=TRUE}

```

```{r unconditional-hint}
#Fill in the blanks with the functions to make a new variable that is the mean of GPA
hsls_clean<-hsls_clean%>%
  _____(mean_gpa_uncond=_____(gpa,na.rm=TRUE))
```

```{r unconditional-solution, exercise.reveal_solution = TRUE}
hsls_clean<-hsls_clean%>%
  mutate(mean_gpa_uncond=mean(gpa,na.rm=TRUE))
```

How far off on average is the mean from a given student's actual GPA? 

```{r rmse, exercise=TRUE, exercise.eval=TRUE}

```

```{r rmse-hint-1}
#Calculate the RMSE 

```

```{r rmse-hint-2}
#Use the RMSE function
rmse_uncond<-hsls_clean%>%____(gpa,mean_gpa_uncond)
rmse_uncond
```

```{r rmse-solution}
rmse_uncond<-hsls_clean%>%rmse(gpa,mean_gpa_uncond)
rmse_uncond
```

## Conditional Mean with One Predictor
We think that there may be a relationship between a student's GPA and if they have taken AP or IB classes. Maybe these students have lower GPAs because these classes are graded more rigorously. Or maybe these students have higher GPAs because students are only allowed to take AP/IB if they have a high GPA to begin with. Generate a table of the conditional mean of the outcome variable by levels of the predictor variable.

```{r conditional1, exercise=TRUE, exercise.eval=TRUE}

```

```{r conditional1-hint-1}
#Use the group_by, summarize, and mean functions
#GPA is the outcome, AP/IB is the predictor variable 
```

```{r conditional1-hint-2}
hsls_clean%>%
  _____(ap_ib_credit)%>%
  _____(mean_gpa_ap=_____(gpa,na.rm=TRUE))%>%
  arrange(mean_gpa_ap)
```

```{r conditional1-solution}
hsls_clean%>%
  group_by(ap_ib_credit)%>%
  summarize(mean_gpa_ap=mean(gpa,na.rm=TRUE))%>%
  arrange(mean_gpa_ap)
```

##Is this a better predictor?
Is the conditional mean a better predictor of GPA than the unconditional mean? 
First, make a prediction for GPA at each level of AP/IB and add that to your dataset. 
```{r conditional_mutate1, exercise=TRUE, exercise.eval=TRUE}

```

```{r conditional_mutate1-hint-1}
#Use the mutate function

```

```{r conditional_mutate1-hint-2}
hsls_clean%>%
  _____(ap_ib_credit)%>%
  _____(mean_gpa_ap=_____(gpa,na.rm=TRUE))%>%
  arrange(mean_gpa_ap)
```

```{r conditional_mutate1-solution}
hsls_clean%>%
  group_by(ap_ib_credit)%>%
  mutate(mean_gpa_ap=mean(gpa,na.rm=TRUE))%>%
  ungroup()
```



Second, compare your RMSEs. 

```{r conditional1_rmse, exercise=TRUE, exercise.eval=TRUE}

```

```{r conditional1_rmse-hint-1}
#use the RMSE function
```

```{r conditional1_rmse-hint-2}
rmse_ap<-hsls_clean%>%____(ap_ib_credit,mean_gpa_ap)
rmse_ap
```

```{r conditional1_rmse-solution}
rmse_ap<-hsls_clean%>%rmse(ap_ib_credit,mean_gpa_ap)
rmse_ap

```


```{r letter-a, echo=FALSE}
question("Is the conditional or unconditional mean a better predictor of student GPA in this case?",
  answer("conditional", correct = TRUE),
  answer("unconditional")
)
```


